---
title: "UC Irvine ISI-BUDS 2023 Day 09: GLM"
author: "Zhaoxia Yu"
date: "`r Sys.Date()`"
geometry: "left=0.5cm,right=2.5cm,top=1cm,bottom=1cm"
output: 
  beamer_presentation:
    theme: "Goettingen"
    colortheme: "whale"
    fonttheme: "structurebold"
    slide_level: 2
header-includes:
  \setbeamertemplate{navigation symbols}{}
  \setbeamertemplate{footline}[frame number]
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Learning Objectives

-   Review of LM \vspace{0.5cm}
-   GLM
    -   Logistic Regression
    -   Poisson Regression
    -   Multinomial Regression \vspace{0.5cm}
-   The assumption of independent observations

# Review of LM

## A Linear Model (LM)

-   Suppose
    $$Y = \beta_0 + x_{1} \times \beta_1+ … + x_{p} \times \beta_p + \epsilon,$$
    where
    -   the regressand $Y$ is the response / outcome / dependent /
        endogenous variable
    -   the regressors $(x_1, \cdots, x_p)$ are the $p$ covariates /
        independent / explanatory variables
    -   the random term $\epsilon$ has a zero mean and variance
        $\sigma^2>0$
    -   the intercept is $\beta_0$, the other $p$ coefficients are
        $\beta_1, \cdots, \beta_p$

## A Linear Model (LM)

-   Consider the $i$th observation:
    $$Y_i = \beta_0 + x_{i1} \times \beta_1+ … + x_{ip} \times \beta_p + \epsilon_i , i =1, …, n$$
-   Basic assumptions
    -   $E(\epsilon_i)=0$, which is equivalent to
        \textcolor{red}{$E(Y_i|X_i)=\beta_0 + x_{i1} \times \beta_1+ … + x_{ip} \times \beta_p$}
    -   $Var(\epsilon_i)=\sigma^2$. Note, this is equivalent to say
        \textcolor{red}{$Var(Y_i|X_i)=\sigma^2$}.
    -   $(\epsilon_1, \cdots, \epsilon_n)$ are mutually independent
-   If $(\epsilon_1, \cdots, \epsilon_n)$ are i.i.d. $N(0, \sigma^2)$,
    we can derive t-tests and F-tests
-   Question: what if the assumptions are violated?

# Logistic Regression

## A Motivating Example of GLM

-   A motivating example: Consider a binary response variable, i.e.,
    $Y_i$ takes values of 0 or 1. \vspace{0.5cm}
-   Is LM a good choice for this problem? \vspace{4cm}

## A Motivating Example of GLM (continued)

-   Consider the Alzheimer data
-   We create a binary variable

\tiny

```{r alzheimer_data}
alzheimer=read.csv("alzheimer_data.csv", header = TRUE)
#dim(alzheimer)
#names(alzheimer)
attach(alzheimer) 
#length(unique(id))
alzh=(diagnosis>0)*1 #"*1" to create a 0-1 variable
```

\normalsize

## A Motivating Example of GLM (continued)

\tiny

```{r fig.pos = 'h', fig.height = 2, fig.width = 3, fig.align = "center"}
alzh.lm = lm(alzh ~ age + female + educ+lhippo + rhippo)
par(mar = c(4, 4, 0.5, 0.5))
plot(alzh, predict(alzh.lm)); abline(h=c(0,1), col=2)
```

\normalsize

## A Motivating Example of GLM (continued)

-   Is alzh.lm a good model for alzh? \vspace{0.5cm}
-   Several assumptions of the LM have been violated, and
-   The predicted values using LM are not between 0 and 1!
-   Let $X_i=(x_{i1}, \cdots, x_{ip})^T$, i.e., the vector of covariates
    for the $i$th subject.
-   Let $\pi_i=E(Y_i|X_i)$, the expected probability. We would like to
    make sure that $\pi_i\in [0,1]$
-   How?

## Logistic Regression

-   Consider the a special transformation of $\pi_i$:
    $$logit(\pi_i)=log\frac{\pi_i}{1-\pi_i}\in (-\infty, \infty)$$
    -   This is the so-called "logit" link!
    -   $\pi_i=E[Y_i|X_i]$: probability of having AD for a subject with
        covariates $X_i$.
    -   $\frac{\pi_i}{1-\pi_i}$: odds
    -   $logit(\pi_i)=log\frac{\pi_i}{1-\pi_i}=log\frac{P(Y_i=1|X_i)}{P(Y_i=0|X_i)}$:
        log-odds!

## Logistic Regression

-   We connect $\pi_i$ and a linear function of the covariates $X_i$ by
    assuming
    $$log\frac{\pi_i}{1-\pi_i} = \beta_0 + x_{i1} \times \beta_1+ … + x_{ip} \times \beta_p$$
-   Essentially, we model the log-odds.
-   But $Y_i$ is a random variable. We need a distribution. A natural
    choice is the Bernoulli distribution
    $$Y_i|X_i \sim Bernoulli(\pi_i)$$
-   pmf, mean, variance:

## Logistic Regression

-   Estimation of is typically conducted by maximizing the corresponding
    likelihood function\
-   How to obtain the likelihood function
    -   $E(Y_i|X_i)=\pi_i=\frac{exp\{\beta_0 + x_{i1} \times \beta_1+ … + x_{ip} \times \beta_p\}}{1+exp\{\beta_0 + x_{i1} \times \beta_1+ … + x_{ip} \times \beta_p\}}$
    -   $f(Y_i|X_i) = \pi_i^{Y_i} (1-\pi_i)^{1-Y_i}$, i.e.,
        -   $f(Y_i|X_i)=\pi_i$ if $Y_i=1$
        -   $f(Y_i|X_i)=1-\pi_i$ if $Y_i=0$
    -   independence: $f(Y|X) = \prod_{i=1}^n f(Y_i|X_i)$
    -   $L(\beta_0, \beta_1, \cdots, \beta_p) = f(Y|X)$

## Logistic Regression for Retrospective Studies

-   In the previous slide we model $E(Y_i|X_i)$. Because $Y_i$ is
    binary, we have $E(Y_i|X_i)=Pr(Y_i=1|X_i)$.
-   Retrospective studies are often considered because a prospective
    study might take many years and is costly.
-   In a retrospective study, subjects are recruited based on their
    disease status. Let $z=1$ denote being sampled and $z=0$ otherwise.
    Let $$\begin{aligned}
    Pr(z=1|y=0) & = p_0\\
    Pr(z=1|y=1) & = p_1
    \end{aligned}
    $$
-   For a retrospective study, the logistic regression models
    $Pr(y=1|z=1, x)$

## Logistic Regression for Retrospective Studies

-   Does this affect the interpretation of the parameters?
-   Let $\theta=Pr(y=1|x)$ and $\phi=Pr(y=1|z=1, x)$. By Bayes' theorem
    and assuming that $z$ does not dependent on $x$, 
\tiny    
    $$
    \begin{aligned}
    \phi &= Pr(y=1|z=1, x) \\
    &= \frac{Pr(z=1|y=1,x)Pr(y=1|x)}
    {Pr(z=1|y=1,x)Pr(y=1|x) + Pr(z=1|y=0,x)Pr(y=0|x)}\\
    &= \frac{p_1\theta}{p_1\theta + p_0 (1-\theta)}
    \end{aligned}
    $$
\normalsize    


## Logistic Regression for Retrospective Studies

-   Therefore $$
    log(\frac{\phi}{1-\phi})=log(\frac{p_1\theta}{p_0(1-\theta)})=log(p_1/p_0) + log(\frac{\theta}{1-\theta})
    $$

-   The result suggests that, when using logistic regression,

    -   the only difference between a prospective study and a
        retrospective study would be the intercept.
    -   the inference for the other parameters is still valid even
        though the subjects were recruited based on their disease status
        (such as a retrospective case-control study)

## Logistic Regression

-   How to obtain the maximum likelihood estimates (MLE) of the
    parameters ($\beta_0, \cdots, \beta_p$)?
    -   Iteratively re-weighted least squares (IRLS): the default method
        used by R
    -   The Newton-Raphson algorithm

## The Motivating Example of Logistic Regression

\tiny

```{r, fig.height=3}
alzh.glm = glm(alzh ~ age + female + educ+lhippo, family=binomial)
par(mar = c(4, 4, 0.5, 0.5))
plot(alzh, predict(alzh.glm, type="response")); abline(h=c(0,1), col=2)
#More visualizaitons
#https://blogs.uoregon.edu/rclub/2016/04/05/plotting-your-logistic-regression-models/
```

\normalsize

## Interpreting a logistic regression

\tiny

```{r}
summary(alzh.glm)$coefficients[-1,]
```

\normalsize


## Interpreting a logistic regression

-   Consider the age variable. The estimated coefficient is 0.018138.
    What information does it provide?
-   The estimated log-odds AD for subject $i$ is (or add a constant
    \textcolor{red}{determined by study design}, see the slides about retrospective studies)
    $$logit(\hat\pi_i) = \hat\beta_0 + \hat\beta_{age} age_i + \hat\beta_2 female_i + \hat\beta_3 educ_i + \hat\beta_4 lhippo_i$$
-   Let $\tilde \pi_i$ denote estimated log-odds after one year
    $$logit(\tilde\pi_i) = \hat\beta_0 + \hat\beta_{age} (age_i+1) + \hat\beta_2 female_i + \hat\beta_3 educ_i + \hat\beta_4 lhippo_i$$

## Interpreting a logistic regression

-   The estimated change in log-odds
    $$logit(\tilde\pi_i) - logit(\hat\pi_i)=log\frac{\tilde \pi_i}{1-\tilde \pi_i} -log \frac{\hat \pi_i}{1-\hat \pi_i}=0.018138$$
-   Take exponential of both sides, we have
    $$\frac{\frac{\tilde \pi_i}{1-\tilde \pi_i}}{\frac{\hat \pi_i}{1-\hat \pi_i}} = exp(0.018138)$$
-   The odds of AD in one year later is
    $exp(0.018138)=$\textcolor{red}{1.018303 times} of the current odds.
-   The estimated increase in odds of AD in a year is
    $e^{0.018138}-1=1.8303\%$

## Interpreting a logistic regression

-   A $95\%$ confidence interval
    -   First, obtain a $95\%$ C.I. for the difference in log-odds:
        $(0.018138-1.96*0.004246, 0.018138+1.96*0.004246)=(0.00982, 0.0265)$
    -   Then, we transform them to increase in odds:
        $(e^{0.00982}-1, e^{0.0265}-1)=(0.99\%, 2.69\%)$

## Interpreting a logistic regression

-   What if we are interested in the increase in odds of AD in ten years
    (everything else is fixed)?
-   The estimated increase in odds of AD in 10 years is
    $$e^{10*0.018138}-1=19.89\%$$
-   A $95\%$ C.I. for 10-year increase in odds:

\tiny

```{r}
exp(10*c(0.018138-1.96*0.004246, 0.018138+1.96*0.004246))-1
```

\normalsize

i.e., $(10.3\%, 30.3\%)$

## Interpreting a logistic regression

-   Very often, we also want to know the significance of a variable
    after adjusting for other important covariates?
-   Does age show a significant effect after adjusting for gender,
    education, and hippocampus volume?
-   A test for $H_0: \beta_{age}=0$ using the Wald test (a type of
    large-sample test)

\tiny

```{r}
summary(alzh.glm)$coefficients["age",]
```

\normalsize

-   Other tests, such as likelihood ratio test, can also be used


# GLM

## GLM

-   Recall that we used the \underline{logit} link in the logistic
    regression $$g(\pi_i)=logit(\pi_i)=\frac{\pi_i}{1-\pi_i},$$ where
    $\pi_i=E(Y_i|X_i)$.
-   How about LM? $g(\mu_i)=\mu_i$, where $\mu_i=E(Y_i|X_i)$. LM uses the \underline{identity} link
-   Poisson $g(\lambda_i)=log(\lambda_i)$, where $\lambda_i=E(Y_i|X_i)$, and $Y_i|X_i \sim Poisson(\lambda_i)$.

# Poisson Regression

## Poisson Regression: The Model

-   Poisson regression is often used to model count data
-   Why are count data special?
    -   Count data are non-negative
    -   Count data take integer values
-   Count data often violate the assumption of \`\`constant variance"
    -   Count data often follow a Poisson distribution
    -   Consider $K\sim Poisson(\lambda)$. $E(K)=?, Var(K)=, pmf$?

## Poisson Regression: Motivating Example

-   Neurons may \underline{fire} selectively for particular types of
    stimuli
-   To understand whether a neuron is a visual-selective neuron, 20
    trials were run for each of the five image categories:
    -   animal, fruit, kids, military, space
-   In each trial, the number of spikes (the number of times that the
    neuron fired) within a 1-second window was recorded

## Poisson Regression

\tiny

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
#https://www.ics.uci.edu/~zhaoxia/Data/chosen_neuron_data.csv
chosen_neuron_data <- read_csv(
  "https://www.ics.uci.edu/~zhaoxia/Data/chosen_neuron_data.csv")
chosen_neuron_data <- chosen_neuron_data[, c(2:4)]
dim(chosen_neuron_data)
names(chosen_neuron_data)
```

\normalsize

## Poisson Regression

```{r}
attach(chosen_neuron_data)
# Even split of image categories among trials
table(image_categ)
sapply(split(n_spikes, image_categ), mean)
```

## Poisson Regression: Visualize the count data (by image category)

```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Plots -------------------------------------------------------------------
library(tidyverse)
library(gridExtra)
library(grid)
image_category_colors <- c("#D55E00", "#009E73", "#E69F00", "#56B4E9", "#CC79A7")

# Spike counts per trial
neuron_trial_plot <- chosen_neuron_data %>% 
  mutate(ordered_trial_num = as.numeric(
    factor(trial_number, levels = unique(trial_number[order(image_categ)]))
    )) %>% 
  ggplot(aes(ordered_trial_num + 0.5, n_spikes,
             fill = image_categ)) +
  geom_col() +
  labs(y = "Total Spikes Detected\n(0.2 to 1.2 seconds after stimulus onset)", 
       x = "Trial (sorted by image category)", 
       fill = "Image Category") +
  scale_fill_manual(values = image_category_colors) +
  scale_y_continuous(breaks = seq(0, 10, by = 1), minor_breaks = NULL) +
  scale_x_continuous(breaks = seq(0, 100, by = 20), minor_breaks = seq(0, 100)) +
  coord_cartesian(expand = FALSE) +
  theme_bw() +
  theme(text = element_text(size = 10),
        legend.position = "bottom")
```

```{r, fig.height=2, fig.width=4}
neuron_trial_plot
```

## Poisson Regression

\tiny

```{r}
# Fit generalized linear model 
# Poisson GLM with the default log link function
poisson_fit <- glm(n_spikes ~ image_categ-1, 
                   data = chosen_neuron_data,
                   family = poisson(link="log"))
# Tabulate the coefficient estimates
poisson_neuron_table <- summary(poisson_fit)$coefficients
row.names(poisson_neuron_table)=c("Animal", "Fruit", 
                                  "Kids", "Military", "Space")
```

\normalsize

## Poission Regression: Model Summary

\tiny

```{r, tidy=TRUE}
poisson_neuron_table
```

\normalsize

## Poisson Regression: Visualize Observed v.s. Fitted

```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Comparison of spike count distribution with Poisson fit
poisson_dist_params <- chosen_neuron_data %>% 
  mutate(image_categ_number = as.numeric(factor(image_categ)) - 3) %>% 
  group_by(image_categ) %>% 
  summarise(predicted_trial_counts = c(sapply(mean(n_spikes), 
                                              function(x) {dpois(0:9, x)}) * 20),
            spike_count = c(0:9) + unique(image_categ_number) * 0.1) %>% 
  ungroup()

# Spike counts per image category (vs Poisson fits)
poisson_obs_fit_plt <- chosen_neuron_data %>% 
  ggplot(aes(n_spikes, fill = image_categ)) +
  geom_histogram(binwidth = 0.5, position = "dodge") +
  geom_point(aes(spike_count, predicted_trial_counts, 
                 color = "Poisson Distribution Fit"),
             data = poisson_dist_params) +
  labs(x = "Total Spikes Detected (0.2 to 1.2 seconds after stimulus onset)", 
       y = "Number of Trials", fill = "Image Category", color = "") +
  scale_fill_manual(values = image_category_colors, 
                    guide = guide_legend(override.aes = list(shape = NA))) +
  scale_color_manual(values = "black") +
  scale_x_continuous(breaks = seq(0, 10), minor_breaks = NULL) +
  scale_y_continuous(breaks = seq(0, 20, by = 4)) +
  coord_cartesian(xlim = c(-0.5, NA), ylim = c(0, 20), expand = FALSE) +
  theme_bw() +
  theme(text = element_text(size = 14),
        legend.position = "bottom")
```

```{r, fig.height=4}
poisson_obs_fit_plt
```

## The Deviance of GLM object

-   Next, we would like to discuss the significance of the image_categ
    variable. To do so, we first look at the deviance of a GLM object
-   The deviance of a GLM object obj is
    $$2[log(L_{saturated})- log (L_{obj})]$$
-   What is the null hypothesis of no visual-selection?
    $$H_0: \beta_{Animal}=\beta_{Fruit}=\beta_{Kids}=\beta_{Military}=\beta_{Space}$$
-   What is the d.f. in a likelihood ratio test (LRT)?

## Poission Regression: The Overall Significance

-   Consider two \underline{nested} models obj1 and obj2, the difference
    in their deviances is $$2[log(L_{obj2}) - log(L_{obj1})],$$ which is
    the LRT statistic.
-   What is the saturated model?
    -   Logistic: $\pi_i=y_i$ and $L_{saturated}=1$
    -   Poisson: $\lambda_i = y_i$ and
        $L_{saturated}= \prod_i \frac{y_i^{y_i}e^{-y_i}}{y_i!}$.

## Poission Regression: The Overall Significance

\tiny

```{r}
# Test for visual selectivity: Likelihood Ratio Test
poisson_fit0 = glm(n_spikes ~1, data=chosen_neuron_data, family=poisson(link="log")) 
anova(poisson_fit0, poisson_fit, test = "LRT")
```

\normalsize

## Poission Regression: The Overall Significance

\tiny

```{r}
# Test for visual selectivity: Rao's score test
anova(poisson_fit0, poisson_fit, test = "Rao")
```

\normalsize

## Poission Regression: The Overall Significance

\tiny

```{r}
# Test for visual selectivity: Wald test. I will explain reparameterization in a few slides
poisson_fit_repara <- glm(n_spikes ~ image_categ, data = chosen_neuron_data,
                   family = poisson(link="log"))
Wald.stat=poisson_fit_repara$coefficients[2:5] %*% 
  solve (summary(poisson_fit_repara)$cov.unscaled[2:5,2:5]) %*% 
  poisson_fit_repara$coefficients[2:5]
1-pchisq(Wald.stat, df=4)
```

\normalsize

## Poission Regression: Model Interpretation

\tiny

```{r}
summary(poisson_fit)$coefficients
```

-   $\hat\beta_{Fruit}=1.2809$: What does it tell us?
-   Recall that we used the log link \vspace{3cm}

## Poission Regression: Model Interpretation

-   Note that the model poisson_fit does not include $\beta_0$.
-   That's why we can estimate $\beta_{Animal}$, $\beta_{Fruit}$,
    $\beta_{Kids}$, $\beta_{Military}$, $\beta_{space}$.
-   Question: how should we interpret the estimated coefficients if the
    intercept term was included?
    -   Try poisson_fit_repara \<- glm(n_spikes \~ image_categ, data =
        chosen_neuron_data, family = poisson(link="log"))
    -   Are the two models equivalent? (Lab activity)

## Poisson Regression: Model Interpretation

-   Re-parameterization

```{r, message=FALSE, echo=FALSE}
poisson_fit_repara <- glm(n_spikes ~ image_categ, data = chosen_neuron_data,
                   family = poisson(link="log"))
```

-   Parameters:
    -   poisson_fit:
    -   poisson_fit_repara:
-   Compare the summary of the two models:
-   Are they different models? \vspace{2cm}

## Parameterization 1: without intercept

\tiny

```{r}
summary(poisson_fit)
```

\normalsize

## Parameterization 2: with intercept

\tiny

```{r}
summary(poisson_fit_repara)
```

\normalsize

## Linear Functions of Parameters

-   The Poisson regression we fit provides estimates of
    $\beta_{Animal}$, $\beta_{Fruit}$, $\beta_{Kids}$,
    $\beta_{Military}$, $\beta_{space}$, which are the log of the
    Poisson rates
-   What if we are interested in difference between specific groups?
    e.g.,
    -   $\beta_{Fruit}-\beta_{Animal}$
    -   $\frac{\beta_{Fruit}+\beta_{Animal}+\beta_{Kids} +\beta_{Military} + \beta_{Space} }{5}$
    -   $\beta_{Fruit}-\frac{\beta_{Animal}+\beta_{Kids} +\beta_{Military} + \beta_{Space} }{4}$
-   They are linear functions of the coefficients, i.e., in the form of
    $a^T \beta$, where $a$ is a 5-by-1 vector.

## Linear Functions of Parameters

-   LM/GLM provides not only estimated coefficients but also the
    variance-covariance of the estimated covariates
    -   Let $\hat\beta=c(\hat\beta_1, \cdots, \hat\beta_p)^T$
    -   Let $\hat \Sigma$ denote the estimated variance-covariance of
        $\hat\beta$
    -   Let $a$ be linear coefficients \vspace{3cm}

## Inference of Linear Functions of Parameters

-   Consider a linear function : $a^T\beta$
-   Estimate: $a^T\hat\beta$
-   Variance of the estimate: $Var(a^T\hat\beta)=a^T\hat\Sigma a$
-   Standard Error (SE): $s.e.(a^T\hat\beta)=\sqrt{a^T\hat\Sigma a}$
-   A $95\%$ confidence interval:
    $$(a^T\hat\beta - 1.96*s.e.(a^T\hat\beta), a^T\hat\beta + 1.96*s.e.(a^T\hat\beta))$$
-   Z-value: $\frac{a^T\hat\beta- a^T\beta}{s.e.(a^T\hat\beta)}$

## Inference of Linear Functions of Parameters: Example

-   Parameter of interest:
    $\frac{\beta_{Fruit}+\beta_{Animal}+\beta_{Kids} +\beta_{Military} + \beta_{Space} }{5}$

\tiny

```{r, message = FALSE, warning = FALSE, echo = TRUE}
a=matrix(rep(1/5,5), 1)
a%*%poisson_fit$coefficients #estimate
sqrt(a%*%summary(poisson_fit)$cov.unscaled%*%t(a)) #s.e.
```

\normalsize

## Linear Contrasts

-   Linear contrasts are a special family of linear functions
    \vspace{0.3cm}
-   We say $a^T\beta=\sum_i a_i \beta_i$ is a linear contrast if
    $\sum a_i=0$, where $a=(a_1, \cdots, a_p)^T$. \vspace{0.3cm}
-   Often, we are interested in whether a linear contrast is zero, i.e.,
    $H_0: a^T \beta=0$ \vspace{0.3cm}
-   z-value: $\frac{a^T\hat\beta - 0 }{s.e.(a^T\hat\beta)}$

## Linear Contrast (e.g., Fruit vs Animal)

\tiny

```{r,  message = FALSE, warning = FALSE, echo = TRUE}
a <- matrix(c(-1, 1, 0, 0, 0), 1)
#estimate
fruit_animal_est = a%*%poisson_fit$coefficients
#variance
fruit_animal_var = a%*%summary(poisson_fit)$cov.unscaled%*%t(a)
#z value
print(fruit_animal_est/sqrt(fruit_animal_var))
```

\normalsize

## Linear Contrast (e.g., Fruit vs Animal) with the multcomp package

\tiny

```{r,  message = FALSE, warning = FALSE, echo = TRUE}
library(multcomp)
a <- matrix(c(-1, 1, 0, 0, 0), 1)
t <- glht(poisson_fit, linfct = a)
summary(t)
```

\normalsize

## Multinomial Logistic Regression

\tiny

```{r, message=FALSE}
library(nnet)
multinom(diagnosis ~ age + female + educ + lhippo + rhippo)
```

\normalsize

## Other concerns

-   Dispersion: under- or over-dispersion
-   Zero-inflated Poisson Regression
-   Model selection ...

\tiny

```{r, fig.height=3}
hist(n_spikes)
#Interested in how to fit a zero-inflated Poisson regression? See the link 
#https://www.rdocumentation.org/packages/pscl/versions/1.5.5/topics/zeroinfl
```

\normalsize

# The Assumption of Independence

## Independent Observations

-   The common assumption we have made in LM and GLM is that the
    observations are independent with each other
-   This is not always the case
-   Examples: \vspace{3cm}

## Independent Observations

-   What is the consequence of ignoring data independence?
    -   The damage is probably worse than violations of distributions
    -   Fortunately, tools have been developed to account for data
        dependence
    -   Linear Mixed-Effects Model (LME)
    -   Generalized Linear Mixed-Effects Model (GLMM)
